\chapter{The Bounce Transform: Arrow, Charge, and Chirality}
\label{chap:chirality}

When matter crosses Cartasis, three things could in principle happen to it: its
\emph{arrow of time} could reverse, its \emph{charge} could flip (matter $\to$
antimatter, the operation $C$), and its \emph{handedness} could be filtered (one
chirality transmitted preferentially). Earlier drafts of this chapter assumed
the bounce performs a $CPT$-like flip that conveniently delivers a
matter--antimatter asymmetry. The dynamics we can actually derive say otherwise,
and the honest result is more constraining---and, in its way, more interesting.

\begin{quote}
\itshape
The minimal Einstein--Cartan bounce is $CPT$-even. It reliably turns a black hole
into a new, expanding universe (Chapter~\ref{chap:sims}), but it does \emph{not}
hand us baryogenesis. Any matter--antimatter asymmetry must be supplied, not
manufactured by the membrane.
\end{quote}

This chapter establishes that statement, then asks what would have to be added to
recover an asymmetry, and which downstream claims (the CMB as parent Hawking
radiation, dark matter as the parent's gravity through the membrane) survive.

\section{What drives the bounce is $C$, $P$, and $T$ even}

In Einstein--Cartan--Sciama--Kibble gravity, torsion is non-dynamical and can be
integrated out algebraically~\citep{hehl1976}. For Dirac matter this leaves a
four-fermion contact term in the effective Lagrangian---the Hehl--Datta
interaction,
\[
  \Delta\mathcal{L}_{\text{tors}}
    = -\,\frac{3}{16}\,\frac{8\pi G}{c^4}\,
      \big(\bar\psi\gamma^5\gamma_a\psi\big)\big(\bar\psi\gamma^5\gamma^a\psi\big),
\]
the (negative) square of the axial current $j^5_a = \bar\psi\gamma^5\gamma_a\psi$.
This is precisely the term whose energy density grows as the square of the spin
density and overwhelms the (linear) gravitational attraction at $\rhoc$: it is
\emph{what makes the bounce} (Chapter~\ref{chap:cartasis}). Its discrete
symmetries are therefore the discrete symmetries of the bounce itself:

\begin{itemize}
  \item \textbf{$C$:} the axial current is charge-conjugation \emph{even} (unlike
    the vector current $\bar\psi\gamma_a\psi$, which is $C$-odd). Its square is
    $C$-even.
  \item \textbf{$P$:} under parity $j^5_0 \to -j^5_0$ and $j^5_i \to +j^5_i$, so
    the contraction $j^5_a j^{5\,a} = (j^5_0)^2 - (j^5_i)^2$ is $P$-\emph{even}.
  \item \textbf{$T$:} the same bilinear-squared structure is $T$-\emph{even}.
\end{itemize}

\paragraph{Result.} The interaction that drives the bounce is separately $C$-,
$P$-, and $T$-even. Three consequences follow immediately and are not
model-dependent:

\begin{enumerate}
  \item \textbf{No flip.} A $C$-even interaction cannot convert matter into
    antimatter. The membrane does not charge-conjugate what passes through it.
  \item \textbf{No filter from nothing.} A $C$- and $P$-even interaction cannot
    \emph{generate} a chiral asymmetry. With zero net axial density in, it gives
    zero net axial density out.
  \item \textbf{No arrow reversal.} A $T$-even dynamics does not reverse the
    direction of proper time (see the next section).
\end{enumerate}

The bounce is real and robust; the convenient asymmetry-generating ``flip'' is
not part of it.

\section{The arrow of time: a Janus point}

In the Tier-1 and Oppenheimer--Snyder bounces (Chapter~\ref{chap:sims}) the
matter worldlines carry a single, monotonic proper time $\tau$: nothing runs
backwards, and the scale factor merely passes through a minimum. There is no
reversal of the geometric time orientation; the bounce is a smooth
\emph{continuation}, collapse $\to$ expansion.

What reverses is the \emph{thermodynamic} arrow. Infalling parent material
arrives from a structured, high-entropy universe; compression toward $\rhoc$
dismantles that structure and smooths the geometry (Weyl curvature $\to 0$), so
gravitational entropy \emph{decreases} toward the membrane. On the far side,
expansion lets structure---and eventually black holes---form again, so entropy
\emph{increases} away from it. Entropy therefore has a \emph{minimum at the
membrane}, and the locally-defined arrow points away from the bounce on both
sides. This is a \emph{Janus point}~\citep{barbour2014}: one spacetime, monotonic
proper time, two thermodynamic arrows diverging from a central entropy minimum.

The baby thus inherits its low-entropy past---the Past Hypothesis of
Chapter~\ref{chap:timearrow}---for free, with \emph{no} time flip required. The
``$r$ becomes timelike, so time inverts'' language is a statement about
Schwarzschild coordinates inside the horizon, not a physical reversal of the
matter's proper time.

\section{The fork in the road}

An entropy-minimum surface admits two geometrically distinct readings, and they
are different physics with different predictions. This is the genuine fork the
framework faces.

\subsection*{Road A --- Continuation (the derived case)}

The two sides are physically distinct, real regions---a real parent interior and
a real baby universe---joined by continuous, $CPT$-even evolution. Time
orientation is preserved, $C$ is preserved (matter stays matter), and the arrow
reflects rather than reverses. \textbf{This is what our simulated bounce is.} Its
price is honesty: the bounce supplies a universe, not an asymmetry.

\subsection*{Road B --- $CPT$ mirror (an added postulate)}

One may instead \emph{postulate} that the membrane is a $CPT$ reflection, so that
the region on the far side is the $CPT$ image of this one---reversed time
orientation, parity-flipped, charge-conjugated, hence an \emph{antimatter}
partner. This is the Boyle--Finn--Turok ``$CPT$-symmetric universe''
construction, in which the Big Bang itself is a $CPT$ mirror~\citep{boyle2018}.
It is a coherent and predictive hypothesis (it implies, for example, a definite
right-handed-neutrino dark-matter candidate), but it is an \emph{additional
discrete-symmetry assumption imposed on the bounce}, not a consequence of
Einstein--Cartan dynamics. The old ``Mechanism A: flip'' was Road~B smuggled in
as though derived.

\medskip
\noindent The framework's baseline is Road~A. Road~B remains available as an
explicit, labelled postulate to be judged on its own predictions---but it should
never again be presented as something the geometry gives for free.

\section{Where an asymmetry could actually come from}

Granting Road~A, the membrane neither creates nor flips an asymmetry. There are
exactly three honest places one could be sourced.

\paragraph{The seed (the load-bearing mechanism).} An OG universe nucleated
directly from a vacuum fluctuation inherits the specific quantum numbers of
\emph{that} fluctuation, not the vacuum average. The chirality/baryon bias of a
particular Cartasis-density fluctuation is drawn from a distribution that, by
$CPT$ invariance of the vacuum, is symmetric overall: equal numbers of
matter-biased and antimatter-biased OG lineages across the supraverse. Our
universe then has a definite bias because it descends from a definite, biased
seed. No baryogenesis mechanism is invoked; the asymmetry is \emph{inherited},
and global $CPT$ is preserved at the supraverse level. This is the only source
that survives in minimal ECSK, and it should carry the weight.

\paragraph{Mean-field amplification (not a source).} In a background that already
carries a net axial density $\langle j^5\rangle \neq 0$, the Hehl--Datta energy
depends on alignment, so a pre-existing bias is self-consistently enhanced near
$\rhoc$. This is a genuine effect, but it is an \emph{amplifier}, not a source:
$\langle j^5\rangle = 0$ in gives $0$ out. It can sharpen an inherited bias; it
cannot start one.

\paragraph{Parity-violating gravity (a derived filter, if you pay for it).} A
true chirality \emph{filter}---preferential transmission of one handedness---
requires a parity-odd ingredient. The minimal Hehl--Datta term has none, but the
gravitational action admits parity-odd extensions (the Holst term with a finite
Barbero--Immirzi parameter, or a Nieh--Yan coupling) that make torsion couple
chirally. Adding one is a specific, testable modification, and it is the honest
route to a calculable filter efficiency to compare against observation. Whether
such a term is warranted is an open question, flagged as speculative.

\section{What survives, and what needs massage}

It is worth being explicit about which downstream claims depend on the discarded
flip and which do not.

\begin{itemize}
  \item \textbf{Black hole $\to$ universe: intact.} The central result---a torsion
    bounce reliably producing an inverse-bubble universe behind the horizon---is
    independent of all of this (Chapter~\ref{chap:sims}). The membrane is a
    reliable cosmogonic engine.
  \item \textbf{CMB $=$ parent Hawking radiation: intact, recast.} Under Road~A the
    radiation crossing in is same-charge thermal radiation; there is no
    chirality-distortion signature to predict, and the old argument that ``flip
    preserves thermality, so observations favour flip'' is withdrawn. The
    identification stands or falls on the spectral and geometric calculation of
    Chapter~\ref{chap:darksector}, not on the transform.
  \item \textbf{Dark matter as the parent's gravity through the membrane: intact in
    principle.} The gravitational influence of parent-side mass-energy on the
    baby does not require a flip. What \emph{does} require massaging is the
    specific ``dark matter $=$ filter-rejected \emph{antimatter}, with
    $f\approx1/6$'' story: in minimal ECSK there is no chirality filter to do the
    rejecting, so either (i) the rejected component is ordinary parent matter held
    at the membrane by the junction conditions (no antimatter needed), or (ii)
    one invokes the parity-violating extension above. State which; do not assume
    the filter.
  \item \textbf{Matter--antimatter asymmetry: needs work.} It is now \emph{inherited
    from the seed}, possibly amplified, rather than generated at the bounce. This
    is more honest but less complete, and it removes the tidy
    ``alternating-chirality lineage'' picture that previously resolved the
    ``hell upward'' problem (Chapter~\ref{chap:supraverse}). That resolution must
    be revisited: with no flip, lineage viability and the avoidance of
    near-symmetric ancestors have to be re-argued from seed statistics and
    standard-model dynamics, not from a bounce-induced sign alternation.
\end{itemize}

\section{Status and open questions}

The reassessment trades a convenient story for a defensible one. What is settled:
the bounce is $CPT$-even; the arrow is a Janus reflection, not a reversal; the
asymmetry is not a free output of the membrane. What is open, and where the
framework needs development:

\begin{itemize}
  \item Quantify the mean-field amplification gain as a function of density and
    inherited bias (amplifier-not-source, made numerical).
  \item Decide whether to adopt a parity-violating gravitational term, and if so
    compute the resulting filter efficiency from first principles.
  \item Re-derive the ``hell upward'' resolution and lineage viability without the
    flip, using seed statistics plus standard-model $CP$ violation.
  \item Decide, as a matter of framework commitment, between Road~A (baseline) and
    Road~B (the $CPT$-mirror postulate), and propagate the choice through
    Chapters~\ref{chap:baby} and~\ref{chap:supraverse}.
\end{itemize}

The working derivations behind this chapter are collected in
\texttt{notes/bounce-transform.md}.
