\chapter{The Scoreboard}
\label{chap:scoreboard}

A worldview earns its keep by explaining more from less, and by being honest about
what it has not yet shown. This final chapter tallies both: a head-to-head against
$\Lambda$CDM and the Standard Model, phenomenon by phenomenon; an explicit account
of what would prove the framework wrong; and a sober statement of where each claim
stands and what it still owes. The single line to carry through all of it is the
one the book is named for. In the standard picture the Big Bang is an event,
finished in the first instant, and the dark sector is a pair of unexplained
ingredients added afterward to make the books balance. Here the Big Bang is a
\emph{surface}, still present at our horizon and still feeding, and the dark sector
is not an addition but the live, measurable proof that our dawn is ongoing. That is
the deepest divergence, and most of the scoreboard is its consequences.

\section{Head to head}

The framework's economy is the heart of the case. Where the standard account
introduces a separate device for each puzzle---an inflaton for flatness, a
$\Lambda$ for acceleration, a particle for the missing mass, a baryogenesis
mechanism for the matter excess, and a fine-tuned low-entropy start by
fiat---Eternal Dawn derives all of them from four axioms and one bounce
(\Cref{tab:scoreboard}).

\begin{table}[ht]
  \centering
  \footnotesize
  \setlength{\tabcolsep}{4pt}%
  \resizebox{\textwidth}{!}{%
  \begin{tabular}{@{}p{0.15\textwidth}p{0.37\textwidth}p{0.40\textwidth}@{}}
    \toprule
    \textbf{Phenomenon} & \textbf{$\Lambda$CDM / Standard Model} & \textbf{Eternal Dawn} \\
    \midrule
    The beginning & A one-time Big Bang, finished in the first instant; a real
      initial singularity & An \emph{ongoing} bounce at our horizon, still
      extruding matter; no singularity (torsion cap) \\
    Low-entropy start & Postulated (the Past Hypothesis) & Forced by the bounce
      (Weyl $\to 0$): a Janus point, low entropy for free \\
    Inflation & An inflaton field with a tuned potential & Torsion-driven
      post-bounce expansion; scale-invariant with no inflaton \\
    Matter excess & Sakharov conditions $+$ leptogenesis (manufactured) &
      Inherited from the founding seed, carried by conserved $B-L$ \\
    Dark matter & A new particle (so far undetected) & The parent's mass felt
      through the membrane---geometric, and ongoing \\
    Dark energy & A cosmological constant $\Lambda$ (eternal, unexplained) & The
      parent's ongoing accretion---a \emph{rate}, hence evolving \\
    CMB & A recombination relic at $z\sim1100$ & The parent's Hawking glow,
      redshifted by our expansion \\
    Arrow of time & An unexplained low-entropy boundary & Manufactured at each
      bounce; the supraverse itself is timeless \\
    Fine-tuning & Anthropics over an unspecified multiverse & A typical viable
      universe on a derived distribution \\
    Hubble tension & Unresolved & Inhomogeneous (timescape) clock rate---no new
      parameter \\
    Evolving dark energy & A tension with $\Lambda$ & Predicted: accretion is a
      rate ($w_0>-1$, $w_a<0$) \\
    Preferred axes & A fluke or a systematic & The parent's rotation axis, shared
      by the CMB and galaxy spin \\
    Ultra-large structure & Awkward for the cosmological principle & Inherited
      lumpiness---the mark of a child, not a first universe \\
    Lepton asymmetry & Assumed comparable to the baryon one & Predicted
      $\xi_\nu\sim\eta$ (sphaleron-locked); a large value forbidden \\
    Boltzmann brains & An eternal-$\Lambda$ future is a brain factory & Finite dark
      energy $+$ an arrowless void: none ever form \\
    Postulates & Inflaton, $\Lambda$, a dark-matter particle, baryogenesis, special
      initial conditions & Four axioms and one bounce \\
    \bottomrule
  \end{tabular}}
  \caption{Eternal Dawn versus the standard account, phenomenon by phenomenon. The
    recurring pattern is one device per puzzle on the left, one mechanism for all of
    them on the right---and, throughout, a beginning that is over versus one that is
    still under way.}
  \label{tab:scoreboard}
\end{table}

Two cautions keep this honest. First, $\Lambda$CDM \emph{works}: it fits the data
it was built for with great precision, and Eternal Dawn must reproduce every one of
those successes before its wider reach counts for anything---which, for the CMB
peaks and the expansion history, it does by running the same post-bounce hot Big
Bang (Chapter~\ref{chap:sims}). Second, several of the right-hand entries are
mechanisms whose \emph{numbers} are computed only to first order---the dark-matter
ratio is narrowed to the right decade but not pinned to three digits, the tensor
sector gives a fork rather than a single $r$, the dark-energy history is still
illustrative. The scoreboard is a map of explanatory scope, not a claim of victory.

\section{Where each claim stands}

The entries sort cleanly by how they will be settled.

\paragraph{Already suggestive (Chapter~\ref{chap:nowobs}).} Evolving dark energy of
the predicted shape (DESI), a Hubble tension resolved with no new parameter, more
ultra-large structure than a smooth start allows, and a doubly-anchored matter
excess are all in hand---each consistent with the framework and a standing tension
for $\Lambda$CDM. The single cleanest way to turn suggestion into evidence is the
\emph{shared axis}: a de-confounded galaxy-spin axis that agrees with the CMB
``axis of evil'' would be a uniquely-ours signature, on archival and near-future
survey data.

\paragraph{The next decade decides (Chapter~\ref{chap:nextobs}).} The primordial
tensors (the bounce ties the tilt to $n_T=n_s-1$, breaking inflation's $n_T=-r/8$;
LiteBIRD), the lepton
asymmetry (the framework forbids the large value the data still allow), the
dark-matter null, and large-scale-structure patterns are the near-future forks.

\paragraph{Computation: first results in hand, precision still owed
(Chapter~\ref{chap:sims}).} Several of these are now computed. The CMB peak
\emph{heights} reproduce the real Planck spectrum at $\chi^2/N\simeq1$; the
dark-to-baryon ratio $f\approx1/6$ falls out naturally as an order-unity
torsion-barrier pass-fraction; the primordial tensors give a clean
consistency-relation fork ($n_T=n_s-1$). What remains is precision and depth: the
peak heights at full inhomogeneous-bounce fidelity, the \emph{exact} $f$ from the
membrane profile rather than its decade, and a rigorous supraverse measure. The
results so far were each bounded at days to weeks; the remaining precision targets
are the longer haul.

\section{What would prove it wrong}

A framework that could not fail would not be science. Eternal Dawn is falsifiable,
and these are the clean ways it dies:
\begin{enumerate}
  \item \textbf{Two cosmic axes that disagree.} If a de-confounded galaxy-spin axis
    and the CMB anomaly axis are both significant but point in \emph{different}
    directions, one parent spin cannot have imprinted both. (A mere \emph{absence}
    of an axis does not falsify it---most universes are low-spin, so a quiet sky is
    typical.)
  \item \textbf{A dark-matter particle.} A confirmed fundamental dark-matter
    particle kills the geometric reading.
  \item \textbf{Dark energy proven constant.} If DESI and successors pin $w=-1$ at
    higher precision, dark-energy-as-accretion fails.
  \item \textbf{A large primordial lepton asymmetry.} A measured $\xi_\nu\gtrsim
    10^{-2}$ has no source in the inherit-then-sphaleron-process picture.
  \item \textbf{Inflationary tensor waves.} Primordial gravitational waves at
    inflation-predicted levels challenge the bounce-as-inflation-substitute.
  \item \textbf{The anomalies dissolve.} If the ``axis of evil'' and its kin are
    proven foreground or statistical artefacts, the parent-rotation signatures are
    simply absent.
  \item \textbf{The bounce cannot reproduce cosmology.} If detailed Einstein--Cartan
    bounce computations cannot return the CMB and structure formation, the framework
    is wrong in detail however appealing its shape.
\end{enumerate}

\section{Where we stand}

Strip the scoreboard to one sentence and it reads: the same physics that forbids a
singularity---continuous, conservative, with torsion capping the collapse---returns,
with almost no further assumption, the bounce, the arrow of time, the matter excess,
the dark sector, the microwave background, and a beginning that never ended. The
Standard Model wrote, with great care, the biography of a parentless first
universe---an OGU, with its finished Big Bang and its empty dark sector---and took
it for our own. We are not that. We are a child, our dawn still breaking at the edge
of the sky, and the dark matter and dark energy that the textbook adds by hand are
the parts of that ongoing dawn we can already measure.

None of this makes the framework correct. It makes it \emph{coherent, parsimonious,
and testable}---a worldview that derives a great deal from very little, stakes
itself on specific predictions, and tells you exactly how to prove it wrong. The
work now is the finding out. The dawn, if the picture is right, is not behind us. It
is still going on.
