\chapter{Ab Initio Genesis}
\label{chap:genesis}

This is the thermal history of matter told in the language of \emph{Eternal Dawn}: not what
happened in the first $10^{-21}$ second of a finished Big Bang, but what happens, still and
continuously, as the extruded field crosses the bounce membrane and cools. ``Ab initio''
twice over---from first principles (no inserted particle content; everything follows from the
one field's phase diagram) and from the beginning (the literal post-membrane origin). And, in
keeping with the book's central thesis: genesis is \emph{ongoing}. The membrane is a
persistent surface at our past horizon, the parent is still extruding field across it, and the
sequence below is not a clock that ran once but a standing gradient in radius and temperature
that the inflowing field climbs down for as long as the parent feeds.

\section{The one substance, cooling}

Everything in Parts~I and~II rests on a single object: the gravity--torsion field, the same
field whose repulsive core makes the bounce. Matter is not a separate ingredient sprinkled on
spacetime; it is that field, knotted into torsitons. So genesis is not the \emph{creation} of
particles from nothing---it is the \emph{condensation} of particles out of a field that was
already there, as it is squeezed across the membrane and allowed to cool. There is one
substance; genesis is its phase diagram. What we call ``the early universe'' is, in this
picture, the hot end of the standing gradient: the layer nearest the membrane where the field
has not yet cooled enough to condense.

\section{The order of operations}

Read the following as a descent in temperature $T$---equivalently, outward from the membrane,
since the field cools with distance from the hot extrusion surface. Each stage is a threshold
the cooling field crosses.

\paragraph{Stage 0 --- at the membrane: the symmetric, massless field.}
Right at the membrane the field is at the bounce density $\rhoc$ and hottest; the chiral
condensate is melted, $v\to0$. By the configurational-mass result
(Section~\ref{sec:configurational-mass}), when the condensate vanishes so does mass: the
field crossing the membrane carries no rest mass---a hot, symmetric, effectively massless
gravity--torsion plasma. The quantities conserved across the membrane are exactly those
Part~I keeps: $P^\mu$, $B\!-\!L$, $Q$, spin---not mass, which has not yet switched on.

\paragraph{Stage 1 --- the condensate forms: mass switches on.}
As the field cools past the chiral scale, the symmetric phase becomes unstable and the vacuum
condenses, $v:0\to f_\pi$. This is the single most important event in the particle genesis and
the moment mass appears in the universe: every torsiton's rest mass is the condensate-bound
field energy, so before this nothing weighs anything and after it everything does. The scale
$f_\pi$ is not inserted---it is generated by dimensional transmutation of the spin-gravity
coupling (Chapter~\ref{chap:weltformel}), the way $\Lambda_{\mathrm{QCD}}$ is generated. The
dawn lights up---the field acquires weight---precisely here.

\paragraph{Stage 2 --- the torsitons crystallise.}
With a condensate to bind against, the field's localized self-bound solutions become stable:
the torsitons of Chapter~\ref{chap:solitons}. The structureless plasma crystallises into
matter---standing waves freezing out of a cooling continuum, as droplets condense from a
cooling vapour. Each torsiton is a quantum of matter; the generations are its internal rungs,
a geometric ladder with a finite cap, so genesis produces three families and stops.

\paragraph{Stage 3 --- colour and confinement turn on.}
The torsitons carry the three-valued Pauli label the algebra forces
(Chapter~\ref{chap:forces}). As the field cools further, the strong channel passes from the
screening, perturbative regime into the confining one: the condensate, acting as a dual
superconductor, expels colour-electric flux into flux tubes. Above this the coloured torsitons
roam in a deconfined plasma; below it they are roped together and only colour singlets
survive.

\paragraph{Stage 4 --- hadrons.}
Once the tubes are on, an isolated coloured torsiton would cost an infinite tube, so the field
arranges its quanta into the singlets the colour algebra allows: mesons (torsiton $+$
anti-torsiton) and baryons (three torsitons, one of each colour). The protons and neutrons of
the universe condense here.

\paragraph{Stage 5 --- the light elements.}
Cooler still, the baryons assemble into the first nuclei (standard big-bang nucleosynthesis),
and the field's genesis hands off to ordinary nuclear and atomic physics. In \emph{Eternal
Dawn} this is the outer, cool layer of the standing gradient---and, unlike in $\Lambda$CDM, it
is continuously re-supplied from within by the ongoing extrusion, not a frozen relic of a
one-time event.

\section{The same diagram as the bounce}

The descent---symmetric/massless $\to$ condensate $\to$ mass $\to$ torsitons $\to$
confinement---is the particle-scale image of the cosmological bounce run as a phase diagram
instead of a trajectory. At the membrane the field is in its melted, symmetric, massless phase
($\rhoc$, $v=0$, the bounce core); as it cools it condenses and acquires weight; the matter we
are made of is the crystallised, cooled phase. The bounce melts matter back to the symmetric
field on the way in, and genesis re-condenses it on the way out: catharsis and genesis are the
two halves of one crossing.

One honest caveat survives, the density question of Chapter~\ref{chap:weltformel} (and the
Appendix): whether the density at which a particle-scale torsiton stabilises equals the
cosmological $\rhoc$, or is related to it by a computable factor, is the quantitative seam
between this chapter and Part~I---stated plainly rather than papered over.
