\chapter{Open Questions}
\label{app:openq}

The framework is internally coherent but has many unresolved questions. This
appendix catalogs them, both for honesty about the framework's current state and
to guide future work.

\section{Foundational physics}

\paragraph{Q1. Precise value of bounce density}
Current estimates place $\rhoc \sim 10^{50}\,\mathrm{kg\,m^{-3}}$, but this
varies by factors of 10--100 depending on coefficient choices in the
Einstein--Cartan Lagrangian. A clean derivation pinning the value would be
valuable.

\paragraph{Q2. Is bounce density universal?}
Does the bounce happen at the same density for different matter content
(different fermion species, different bound states)? Or is the bounce condition
matter-dependent?

\paragraph{Q3. How does the bounce interact with rotation?}
Most analyses assume spherical (Schwarzschild) bounces. Real astrophysical black
holes are typically rotating (Kerr). The Kerr--Cartan bounce dynamics may differ
qualitatively from Schwarzschild--Cartan and might preferentially produce
specific chirality dynamics. Detailed analysis is needed.

\paragraph{Q4. Quantum gravity at the bounce density}
The framework treats Einstein--Cartan classically. At
$\rhoc \sim 10^{50}\,\mathrm{kg\,m^{-3}}$, full quantum-gravity effects should be
subdominant (still 46 orders of magnitude below Planck density), but they might
matter for fine details. The full quantum-gravitational picture is unknown.

\paragraph{Q5. Baryon-number conservation}
At trans-electroweak temperatures, sphaleron processes violate baryon number. At
higher temperatures (approaching the bounce density), more exotic violations might occur.
How conserved is $B$ across the bounce? This affects both the baryon-asymmetry
mechanism and the dark-matter interpretation.

\paragraph{Q6. Beyond-Standard-Model physics}
The Standard Model is presumably an effective theory valid up to some scale.
Above that scale, new physics (SUSY? extra dimensions? unknown new fields?) might
dominate. Whether Eternal Dawn is robust to such new physics, or whether it requires the
SM to be the full story up to the bounce density, is unclear.

\section{Bounce dynamics}

\paragraph{Q7. The extrusion coefficient $C$ and the remnant fraction}
The minimal Einstein--Cartan bounce is CPT-even: it neither flips charge nor
filters a pre-existing population (Chapter~\ref{chap:dawn}), so the matter excess
is supplied, not manufactured at the membrane. The live quantitative questions
are then: (i)~the anomaly\,$+$\,sphaleron coefficient $C$ in the
chiral-vortical extrusion estimate $\eta\sim C(\omega/T)$, which must reach
$\sim41$ at maximal spin to source the observed asymmetry at the OGU
(Section~\ref{sec:matter-choice}); and (ii)~the fraction of inherited net $B{-}L$
that survives the extrusion versus is washed out or diluted
(Section~\ref{sec:matter-choice}, and Q16a). These set the asymmetry magnitude and decide
whether lineages are chirally pure or mixed.

\paragraph{Q8. Bounce efficiency}
What fraction of the parent's mass--energy is transmitted to the baby's expanding
phase? Some might remain in an intermediate region, some might be lost to
gravitational waves, some might emerge as Hawking radiation in the parent's
exterior. The efficiency factor is currently a free parameter.

\paragraph{Q9. Information flow through the bounce}
We have argued that information is preserved globally through entanglement
structures across the membrane. The specific quantitative details---what gets
entangled with what, how the Page curve is realized in bounce cosmology---are not
worked out.

\paragraph{Q10. Persistent channel vs.\ severed connection}
We favor the position that the channel persists after the first bounce over the
position that each bounce severs it. But the detailed geometry of a persistent
channel that maintains across the parent black hole's lifetime needs rigorous
treatment.

\section{Supraverse structure}

\paragraph{Q11. Equilibrium establishment}
We argue the supraverse is in thermodynamic equilibrium. How long does
establishment take? Is the supraverse currently in equilibrium or still
approaching it? What sets the equilibrium timescale?

\paragraph{Q12. OG bounce rate}
What is the rate of OG bounce nucleation per unit supraverse spacetime? This
determines whether the supraverse contains many universes or few, and the
dependence on supraverse parameters.

\paragraph{Q13. Universe size distribution shape}
We conjecture the distribution peaks near our observed size. What is the actual
shape? How wide is the distribution? Are most observers in universes very close
to our parameters, or is the distribution wide enough that observers could exist
in significantly different universes?

\paragraph{Q14. Lineage interaction}
Do separate lineages (descended from different OG bounces) ever interact? If so,
what happens at boundaries? Can matter-lineage and antimatter-lineage universes
be near each other in the supraverse?

\paragraph{Q15. Initial conditions for the supraverse}
The framework requires the supraverse to be old enough for equilibrium
establishment, but does not specify what the supraverse ``started from.'' Was
there an initial state, or is the supraverse genuinely eternal? This is partly a
question about the meaning of time at the supraverse level.

\paragraph{Q15a. Do OG universes grow forever?}
Black holes have negative heat capacity, so above
$M_{\mathrm{crit}}=\hbar c^3/(8\pi G k_B T_{\mathrm{bath}})$ a universe-hole
accretes faster than it evaporates and grows. The growth-rate model of
\Cref{sec:shape} gives a partial answer: the runaway is fast
($\dot M\sim M^2$, finite-time blow-up), so slow cooling cannot regulate it, but
\emph{depletion} of the local causal patch does---the hole eats its patch, the
source shuts off, and the mass freezes at $M_{\mathrm{final}}\sim$ the patch mass.
So individual holes are finite, set by their environment, not unbounded. What
remains open is the patch size itself (hence $M_{\mathrm{final}}$) and the void
bath temperature $T_{\mathrm{bath}}$ that fixes the growth time
$\tau_{\mathrm{gen}}\propto T_{\mathrm{bath}}^{-3}$; the latter dials our
generation depth (\Cref{sec:shape}) and is entangled with the measure
problem below and with Q15c.

\paragraph{Q15b. The measure over infinitely many OG universes}
Axiom~\ref{ax:qm} plus infinity implies infinitely many OG nucleations. Making
``we are typical'' meaningful requires a regulated measure over that infinite,
possibly ever-growing ensemble---the measure problem shared with all multiverse
frameworks, here sharpened by ongoing growth (Q15a). Unsolved.

\paragraph{Q15c. The cosmological constant in the void}
An infinite, eternally active quantum vacuum makes acute the question of why the
void is nearly flat rather than de Sitter at the QFT cutoff. The dark-energy
\emph{evolution} is reattributed to parent accretion (Chapter~\ref{chap:home}),
but the bare vacuum-energy problem is untouched, arguably worsened.

\section{Observational connection}

\paragraph{Q16. CMB acoustic peaks}
$\Lambda$CDM predicts specific acoustic peaks in the CMB power spectrum that
match observations beautifully. Can Eternal Dawn reproduce these from bounce dynamics?
This is the central calculational test.

\paragraph{Q16a. Does $\eta$ measure annihilation ash or Hawking inflow?}
\label{q:eta-denominator}
The baryon-to-photon ratio $\eta=n_B/n_\gamma\simeq6\times10^{-10}$ has two
readings of its \emph{denominator}, and they imply different baryogenesis
bookkeeping. (i)~\emph{Annihilation ash}: the photons are the baby's own thermal
entropy, so $\eta$ directly measures the extrusion's net-baryon-per-entropy, and
the maximal-spin chiral-vortical estimate undershoots by $\sim41$
(Chapter~\ref{chap:dawn}, Section~\ref{sec:matter-choice}). (ii)~\emph{Hawking
inflow}: if the CMB is parent Hawking radiation
(Chapter~\ref{chap:home}, Section~\ref{sec:cmb}), the photon
denominator is largely \emph{external}, the baby's internal asymmetry is decoupled
from the observed $\eta$, and the small measured value is partly \emph{dilution} by
the Hawking flood---which can dissolve the factor of $41$ without a large
coefficient $C$. The \emph{numerator} carries the mirror-image worry: a near-pure
matter parent could inject a net $B{-}L$ far above $1\,\mathrm{ppb}$, so matching
$\eta\approx0.6\,\mathrm{ppb}$ requires either inheritance of the parent's own
$\sim$1\,ppb ratio (entropy-conserving extrusion) or dilution down to it
(Section~\ref{sec:matter-choice}). The dilution half of this is now largely answered:
the extruder entropy budget (Section~\ref{sec:extruder}), in both its homogeneous
and shear-dominated channels, gives $D=1$ to $\sim$1 part in $10^{11}$, so the
extrusion is adiabatic and $\eta_{\mathrm{baby}}=\eta_{\mathrm{parent}}$. What
remains is (a)~gravitational particle production at the bounce, the one entropy
source not yet in the budget, and (b)~the genuinely open question---the founder's
\emph{value}, why the original-generation seed sits at $\sim6\times10^{-10}$
rather than $10^{-3}$ or $10^{-15}$, which inheritance propagates but does not
explain. It is entangled with Q5 (baryon-number conservation) and Q8 (bounce
efficiency / entropy budget).

\paragraph{Q17. Hubble tension resolution}
We claim Eternal Dawn might resolve the Hubble tension through modified early-universe
dynamics. The actual calculation showing this has not been done. If it cannot be
done, the framework needs revision.

\paragraph{Q18. Dark-matter density profile}
Observed dark matter has specific clustering properties (NFW profiles,
mass--velocity relations). Can these be derived from parent gravitational
influence through the bounce membrane? Or do they imply something else?

\paragraph{Q19. Galaxy-rotation asymmetry consistency}
If parent rotation imprints galaxy-rotation asymmetry, the relationship should be
specific and predictable. Shamir's observed asymmetry should match a particular
parent rotation axis. Whether this consistency holds at higher statistical
precision is unknown.

\paragraph{Q20. Specific BBN constraints}
Big Bang Nucleosynthesis predicts light-element abundances. Because Eternal Dawn\ shares the
post-bounce radiation-dominated hot Big Bang with $\Lambda$CDM, BBN runs as usual, and
its value of $\eta$ \emph{agrees with the CMB value to $\sim$1\%}: deuterium
$10^5\,\mathrm{D/H}\simeq2.53$ gives $\eta_{10}\simeq6.1$, matching the CMB-fixed
$\omega_b$ (Chapter~\ref{chap:nowobs}, Section~\ref{sec:spin-local}). This is a
concordance, not a tension. What remains open is whether the \emph{approach} to the
radiation era from the bounce (any entropy injection or expansion-rate change near
$\rhoc$) perturbs the neutron freeze-out or the expansion rate at $z\sim10^9$ at the
percent level---a quantitative check the current treatment has not done.

\section{Theoretical issues}

\paragraph{Q21. Connection to string theory / loop quantum gravity}
How does Eternal Dawn relate to other proposed extensions of GR? Could string theory or
LQG be ways of realizing the bounce in different formal frameworks? Are they
competitors or complements?

\paragraph{Q22. Conformal cyclic cosmology relationship}
Penrose's CCC is also a non-singular cyclic cosmology~\citep{penrose2010}. The
recursive cycle of \Cref{sec:recursive-cycle} gives a concrete relationship: a
heat-dead universe's smooth, conformally-flat de Sitter far future is a
low-Weyl-entropy void, which in Eternal Dawn\ nucleates a \emph{scatter} of fresh OGUs rather
than handing off to a single conformally-rescaled successor. Eternal Dawn\ is, on this
reading, ``CCC made to branch.'' What remains open is whether the conformal
matching at the dead-aeon/new-bounce boundary is as clean as CCC's---in particular
whether the de Sitter nucleation rate (the birth rate $\beta$) and the Weyl-entropy
reset can be made quantitative rather than schematic.

\paragraph{Q23. ER\,$=$\,EPR specific realization}
The bounce structure suggests a geometric realization of ER\,$=$\,EPR. What is
the precise mathematical statement? Can it be cleanly derived?

\paragraph{Q24. Holographic principle in the supraverse}
The holographic principle suggests that bulk physics can be encoded on
boundaries. Does the supraverse have a holographic structure? What is the
appropriate boundary? Is each bounce membrane a holographic surface for its
enclosed baby?

\paragraph{Q25. Anthropic considerations}
We have implicitly used anthropic reasoning (observers exist in viable
universes). Can this be made more precise? Does the framework predict where
observers are in the supraverse population in a quantitatively testable way?

\section{Philosophical / interpretive}

\paragraph{Q26. What is the supraverse ``embedded in''?}
Mathematically, the supraverse is a 4D manifold with non-trivial topology.
Physically, what is its ontological status? Is it the totality of existence, or
is there a meta-supraverse containing it?

\paragraph{Q27. Connection between local quantum measurements and bounce
  dynamics}
We argued that gravitational decoherence is the small-scale version of bounce
dynamics. The precise relationship needs working out---is every quantum
measurement literally a miniature bounce-like splitting? Or just analogous?

\paragraph{Q28. Implications for the meaning of physical constants}
If we are typical inhabitants of a supraverse population, our physical constants
are typical for our class of universe. Does this mean physical constants are
statistical features of the supraverse distribution, or are they fundamental? Or
is the distinction meaningless?

\paragraph{Q29. Time at the supraverse level}
The supraverse is static (a block universe / timeless equilibrium), yet it
``supports'' nucleation and equilibrium establishment, which sound dynamical. The
resolution (Chapter~\ref{chap:void}, Section~\ref{sec:no-arrow}): the
supraverse needs \emph{no} arrow---an arrow is a boundary condition, not a law, and the
void has no low-entropy edge. ``What is the stable state of the void?'' is a timeless,
equilibrium question; the supraverse \emph{is} that stationary state, and ``nucleation'' is
the statement that the equilibrium ensemble \emph{contains} fluctuation-condensations,
not that the void evolves toward equilibrium in some external time. The
``dynamical-sounding'' language is an artifact of describing a timeless structure from
inside one of its internal-time condensations. What remains genuinely open is the
deeper interpretive question of whether the supraverse-level ``timelessness'' should be
read as a Wheeler--DeWitt-style frozen formalism or as a Barbour-style timeless
configuration space---a question Eternal Dawn\ shares with all of quantum gravity, and does not
need to settle to be predictive.

\paragraph{Q29a. Boltzmann brains and the simulation worry (largely resolved)}
On a naive de Sitter nucleation count, a Boltzmann brain ($\sim 10^{-10^{69}}$) is
cheaper than an OG seed ($\sim 10^{-10^{84}}$), which would make typical observers
fluctuations---and a cosmology whose only origin is a hand-tuned low-entropy Big Bang
($\sim 10^{-10^{123}}$, \citealp{penrose2010}) is structurally the simulation
hypothesis. The resolution (Chapter~\ref{chap:void},
Section~\ref{sec:not-brains}): a brain---or a simulator---is an information processor
and \emph{requires} a thermodynamic arrow, which the stationary, arrowless,
maximum-entropy void cannot supply~\citep{boddy2014}. Nucleation from the eigenstate
yields only \emph{generic} high-entropy blobs (viable OG seeds, which make their arrow
at the forced low-entropy bounce), never the low-entropy \emph{organized} state of a
brain or a finished Big Bang. Interior brains are then starved by Eternal Dawn's \emph{finite}
dark energy: the recurrence time to fluctuate one ($\sim 10^{10^{69}}$ yr) vastly
exceeds the parent-evaporation cutoff ($\sim 10^{136}$ yr), whereas eternal-$\Lambda$
$\Lambda$CDM has no such cutoff. \emph{What remains open} is a rigorous de Sitter
measure: the qualitative asymmetry (blobs nucleate, minds presuppose an arrow) is
robust, but ``whether a stationary vacuum fluctuates'' is unsettled across quantum
gravity, and a fully quantitative Eternal Dawn\ observer measure does not yet exist.
Reproducible as \texttt{sims/src/cartasis\_sims/boltzmann\_brains.py}.

\section{Methodological}

\paragraph{Q30. Falsifiability completeness}
The framework makes predictions (Chapter~\ref{chap:nowobs}). Are these
collectively sufficient to falsify the framework if it is wrong? Or could the
framework be twisted to accommodate any observational result? A genuinely
scientific framework should be unambiguously falsifiable. We have argued Eternal Dawn is,
but this should be checked carefully.

\paragraph{Q31. Naturalness arguments}
We have argued Eternal Dawn is ``natural'' because it requires few postulates. But
naturalness arguments have led physics astray before (e.g.\ naturalness arguments
for SUSY have not worked out). Is Eternal Dawn's parsimony actually evidence for its
truth, or just for its appeal?

\paragraph{Q32. Relationship to existing alternative cosmologies}
Many alternatives to $\Lambda$CDM exist: timescape (Wiltshire), inhomogeneous
cosmology, modified gravity (MOND/MOG), cyclic universes. Eternal Dawn shares features
with some of these. Detailed comparison would clarify what Eternal Dawn uniquely
contributes and what it borrows.

\section{Practical / programmatic}

\paragraph{Q33. Publication strategy}
The danger is well understood: a single sweeping ``theory of everything'' dropped
whole into the literature by an outsider is reflexively ignored, however careful, and
deservedly so when it cannot be checked piecewise. The antidote is to \emph{not}
publish the synthesis first. The framework's strength is that it factors into
self-contained, individually-checkable results, and those are the publishable units.
A staged plan:

\begin{enumerate}
  \item \textbf{Lead with the narrowest, most defensible quantitative result}, framed
    in the host community's own language, \emph{not} as ``a new cosmology.'' The
    strongest candidate is the de-confounded galaxy-spin/CMB \emph{shared-axis} test
    (Chapter~\ref{chap:nowobs})---a self-contained statistics-on-archival-data paper
    that stands on its own merit whatever one thinks of the framework, and names in
    advance what would confirm or falsify a parent-spin imprint. A second candidate is
    the dark-energy-as-parent-accretion confrontation with DESI ($w_0$--$w_a$ in a
    predicted wedge), again a falsifiable data paper.
  \item \textbf{Then the mechanism papers}, each in the appropriate specialist venue
    and each citing only what it needs: the Einstein--Cartan bounce and its primordial
    spectrum (\emph{Phys.\ Rev.\ D} / \emph{JCAP}); the adiabatic-extrusion
    baryon-inheritance result; the supraverse population statistics and generation
    depth. None of these requires the reader to accept the whole edifice.
  \item \textbf{Preprint everything on arXiv} (gr-qc, astro-ph.CO) with the full
    open-source repository attached, so every figure and number is reproducible from a
    command. Reproducibility is the outsider's single best credential.
  \item \textbf{The monograph last}, as the synthesis for those already persuaded by
    the pieces---this book---explicitly labelled a research program, not a settled
    theory, with epistemic status flagged result by result.
\end{enumerate}
The order matters more than the venue: establish falsifiable pieces that professionals
can check and reuse \emph{before} asking anyone to consider the whole. Honesty about
what is established versus speculative (the discipline this book already imposes) is
the other half of being taken seriously.

\paragraph{Q33a. What \emph{not} to do}
Equally important, given how cosmology-adjacent ``theories'' usually fail: do not lead
with popular-science framing or press; do not claim the framework is proven; do not
present the synthesis to a general-physics audience before the pieces are refereed; do
not overstate any single test (the shared-axis test, e.g., is confirmatory and only
weakly disconfirmatory---Chapter~\ref{chap:nowobs}). The CERN ``black hole eating the
Earth'' episode is the cautionary tale: a misframed claim, amplified by journalists,
costs years of credibility that careful work then has to buy back. Let the
peer-reviewed pieces, and the reproducible code, do the talking.

\paragraph{Q34. Collaboration approach}
The framework benefits from multiple expertises: numerical relativity, cosmology,
quantum field theory, philosophy of physics. The realistic path is to recruit one
collaborator per \emph{piece} rather than a believer in the whole---a
large-scale-structure statistician for the shared-axis paper, an Einstein--Cartan
bounce specialist for the spectrum, a DESI-adjacent cosmologist for the dark-energy
confrontation. Each engages with a self-contained, falsifiable result in their own
field; the synthesis is the author's to carry until the pieces have stood up
independently.

\paragraph{Q35. Funding considerations}
This is theoretical/computational work that does not require large facilities.
Modest computational resources, person-time. What funding mechanisms are
appropriate?

\section{Priority ordering}

The most important open questions for near-term progress:

\begin{enumerate}
  \item \textbf{Q16 (CMB acoustic peaks):} decisive test; must be addressed for
    the framework to be taken seriously.
  \item \textbf{Q7 (flip vs.\ filter):} resolves chirality dynamics, enables the
    dark-matter prediction.
  \item \textbf{Q1 (precise $\rhoc$):} enables quantitative bounce calculations.
  \item \textbf{Q12--Q13 (supraverse statistics):} tests observer-placement
    predictions.
  \item \textbf{Q17 (Hubble tension):} high-impact if it works.
\end{enumerate}

Items 1--3 are calculable now with current Einstein--Cartan understanding plus
standard numerical methods. Items 4--5 require more theoretical development.
